
---
title: "Simulation Challenge"
subtitle: "Generative Models and Monte Carlo Simulation"
format:
  html: default
  pdf: default
execute:
  echo: true
  eval: true
---

# 🎲 Simulation Challenge - Monte Carlo Analysis

## Challenge Overview

**Your Mission:** Create a comprehensive Quarto document that simulates one or two investment strategies, analyzes the results, and demonstrates your ability to present counter-intuitive findings compellingly. Then render the document to HTML and deploy it via GitHub Pages from a new repository called "simulationChallenge."

::: {.callout-warning}
## ⚠️ AI Partnership Required

This challenge pushes boundaries intentionally. You'll tackle problems that normally require weeks of study, but with Cursor AI as your partner (and your brain keeping it honest), you can accomplish more than you thought possible.

**The new reality:** The four stages of competence are Ignorance → Awareness → Learning → Mastery. AI lets us produce Mastery-level work while operating primarily in the Awareness stage. I focus on awareness training, you leverage AI for execution, and together we create outputs that used to require years of dedicated study.
:::

## The Investment Game 🎯

### Original Game Strategy

::: {#exm-ErgodicityEconomicsExample}
Imagine you are offered the following game and given a $1,000 budget in a special account to play the game: I will flip a coin, and if it comes up heads, we increase your account's balance by 50%; if it comes up tails, we reduce your account's balance by 40%. We are not only doing this once, but we will do it once per year until you turn 55. When you turn 55, you will receive the balance in your account.
:::



## Challenge Requirements 📋

### Minimum Requirements for Any Points on Challenge

1. **Create a Quarto Document:** Write a concise quarto markdown file that includes a narrative of what you are doing along with the requested code, results, and visualizations of your simulations.

   ::: {.callout-tip}
   ## 💡 Pro Tip: Source File Reference
   
   **Need help with Quarto syntax or DAFT diagrams?** You can find the complete source `.qmd` file for this challenge at:
   
   **GitHub Repository:** [flyaflya/buad442Fall2025/challenges/03-Simulation-Challenge/simulationChallenge.qmd](https://github.com/flyaflya/buad442Fall2025/blob/main/challenges/03-Simulation-Challenge/simulationChallenge.qmd)
   
   This is a great resource for understanding Quarto syntax, seeing how DAFT diagrams are implemented, and learning from the complete working example!
   :::

2. **Render to HTML:** You must render the quarto markdown file to HTML.

3. **GitHub Repository:** The rendered HTML must be uploaded to a new GitHub repository called "simulationChallenge" in your Github account.

4. **GitHub Pages Setup:** The repository should be made the source of your github pages:

   - Go to your repository settings (click the "Settings" tab in your GitHub repository)
   - Scroll down to the "Pages" section in the left sidebar
   - Under "Source", select "Deploy from a branch"
   - Choose "main" branch and "/ (root)" folder
   - Click "Save"
   - Your site will be available at: `https://[your-username].github.io/simulationChallenge/`
   - **Note:** It may take a few minutes for the site to become available after enabling Pages

## Grading Rubric 🎓

### Questions to Answer for 75% Grade on Challenge

1. **Expected Value Analysis:** What is the "expected value" of your account balance after 1 coin flip for the original game?

2. **Expectation vs. Reality:** Is the expected value positive or negative? Do you expect your account to be worth more or less than $1,000 based on this result?

3. **Single Simulation:** Run one simulation showing the dynamics of your account balance over time. Make an object-oriented matplotlib OR ggplot2 plot showing your simulated account balance over time (i.e. as you age). Comment on the results, are you happy?

### Questions to Answer for 85% Grade on Challenge

4. **Multiple Simulations:** Run 100 simulations modelling the dynamics of your account balance over time. Make an object-oriented matplotlib OR ggplot2 plot showing a probability distribution of the 100 simulatedaccount balance at age 55. Comment on the results, are you happy? Why or why not?

### Questions to Answer for 95% Grade on Challenge

5. **Probability Analysis:** Based on the 100 simulations above, what is the probability that your account balance will be greater than $1,000 at age 55?

### Questions to Answer for 100% Grade on Challenge

6. **Strategy Comparison:** Run 100 simulations for the modified game strategy shown below in @exm-ErgodicityEconomicsExampleModified. What is the probability that your account balance will be greater than $10,000 at age 55? Is this probability higher or lower than the probability in the original game?

### Modified Game Strategy

::: {#exm-ErgodicityEconomicsExampleModified}
Imagine you are offered the following game and given a $1,000 budget in a special account to play the game: I will flip a coin, and if it comes up heads, we increase your bet by 50%; if it comes up tails, we reduce your bet by 40%. You must bet exactly 50% of your current account balance on each flip, and this 50% is locked in for each round. We are not only doing this once, but we will do it once per year until you turn 55. When you turn 55, you will receive the balance in your account.
:::

## Technical Implementation Preferences 💡

### Setting Up Your Analysis

**For R Users:**

- Use `tidyverse` for data manipulation
- Use `ggplot2` for visualizations
- Use `set.seed()` for reproducible results

**For Python Users:**

- Use `numpy` for numerical operations
- Use `pandas` for data manipulation
- Use `matplotlib` (object-oriented)
- Use `np.random.seed()` for reproducible results

### Visualization Preferences

- **Professional Styling:** Use consistent colors, clear labels, readable fonts, and informative titles

## Submission Checklist ✅

**Minimum Requirements (Required for Any Points):**

- [ ] Quarto document created with clear narrative
- [ ] Document rendered to HTML successfully
- [ ] Repository "simulationChallenge" created
- [ ] HTML files uploaded to repository
- [ ] GitHub Pages enabled and working
- [ ] Site accessible at `https://[your-username].github.io/simulationChallenge/`

**75% Grade Requirements:**

- [ ] Expected value calculations shown (Question 1)
- [ ] Expectation vs. reality analysis (Question 2)
- [ ] Single simulation with time series plot (Question 3)
- [ ] Clear interpretation of single simulation results

**85% Grade Requirements:**

- [ ] 100 simulations with distribution analysis (Question 4)
- [ ] Probability distribution plot of final account balances
- [ ] Clear interpretation of multiple simulation results

**95% Grade Requirements:**

- [ ] Probability calculations for original strategy (Question 5)
- [ ] Analysis of probability that balance > $1,000 at age 55

**100% Grade Requirements:**

- [ ] 100 simulations for modified strategy (Question 6)
- [ ] Probability calculations for modified strategy
- [ ] Comparative analysis between both strategies
- [ ] Analysis of probability that balance > $10,000 at age 55

**Code Quality (All Grades):**

- [ ] Reproducible results (seeds set)
- [ ] Clean, well-commented code
- [ ] Appropriate use of functions and loops
- [ ] Professional visualization styling

### Resources

- **Quarto Markdown:** [quarto.org/docs/authoring/markdown-basics.html](https://quarto.org/docs/authoring/markdown-basics.html)
- **Quarto Documentation:** [quarto.org/docs](https://quarto.org/docs)
- **R for Data Science:** [r4ds.had.co.nz](https://r4ds.had.co.nz)
- **Python Data Science Handbook:** [jakevdp.github.io/PythonDataScienceHandbook](https://jakevdp.github.io/PythonDataScienceHandbook)

### Getting Started Tips

::: {.callout-note}
## 🎯 Navy SEALs Motto

> "Slow is Smooth and Smooth is Fast"

*Take your time to understand the simulation mechanics, plan your approach carefully, and execute with precision. Rushing through this challenge will only lead to errors and confusion.*
:::

- **Browse [Essential Simulation Concepts](@sec-simulation-concepts):** This section will give you a good understanding of the concepts you need to know to complete the challenge.
- **Start Simple:** Begin with a single simulation to understand the mechanics
- **Document Everything:** Explain your reasoning and interpret your results
- **Forgetting to Set Seeds:** Always set random seeds for reproducible results
- **Total time to complete:** ~3-4 hours for the 100% grade ⏱️
- **Good luck, and remember simulation will steer you right even when intuition will steer you wrong!** 🎲

::: {.callout-warning}
## 💾 Important: Save Your Work Frequently!

**Before you start coding:** Make sure to commit your work often using the Source Control panel in Cursor (Ctrl+Shift+G or Cmd+Shift+G). This prevents the AI from overwriting your progress and ensures you don't lose your work.

**Commit after each major step:**

- After completing each simulation example
- After finishing each challenge question
- Before asking the AI for help with new code

**How to commit:**

1. Open Source Control panel (Ctrl+Shift+G)
2. Stage your changes (+ button)
3. Write a descriptive commit message
4. Click the checkmark to commit

*Remember: Frequent commits are your safety net!*
:::

## Essential Simulation Concepts 🎯 {#sec-simulation-concepts}

Before diving into the challenge, let's review the key simulation concepts you'll need. These examples will prepare you for the investment game analysis.

### 1. Simple Simulation: Coin Flip Game

Let's start with a basic coin flip simulation to understand the mechanics:


**Key Difference from Investment Game:** Unlike the investment game which models wealth evolution over multiple time periods with multiplicative changes, this simple coin flip game represents a single-period game with additive winnings. The investment game shows how wealth compounds over time ($W_t = 1.5 \times W_{t-1}$ or $W_t = 0.6 \times W_{t-1}$), while this simple game shows fixed winnings ($W = +100$ or $W = -100$) based on a single coin flip outcome.

::: {.panel-tabset}

### R

```{r}
#| label: simple-sim-r
#| fig-cap: R simulation of coin flip game
#| echo: true

suppressPackageStartupMessages(library(tidyverse))

# Set seed for reproducibility
set.seed(123)

# Number of simulations
n_sims <- 10

# Create simulation data
sim_data <- tibble(
  sim_num = 1:n_sims,
  coin_flip = rbinom(n = n_sims, size = 1, prob = 0.5)
) %>%
  mutate(
    winnings = if_else(coin_flip == 1, 100, -100)
  )

# Display results
sim_data
```

### Python

```{python}
#| label: simple-sim-python
#| fig-cap: Python simulation of coin flip game
#| echo: true
#| eval: false

import numpy as np
import pandas as pd

# Set seed for reproducibility
np.random.seed(123)

# Number of simulations
n_sims = 10

# Step 1: Draw coin flips (stochastic node)
X = np.random.binomial(n=1, p=0.5, size=n_sims)

# Step 2: Compute winnings (deterministic node)
W = np.where(X == 1, 100, -100)

# Combine into data frame
sim_data = pd.DataFrame({
    'sim_num': range(1, n_sims + 1),
    'coin_flip': X,
    'winnings': W
})

# Display results
sim_data
```

:::

### 2. Time-Series Simulation: Account Balance Over Time

Now let's simulate how an account balance changes over multiple periods:


**Key Difference from Simple Coin Flip Game:** Unlike the simple coin flip game which represents a single-period game, this time-series model shows sequential balance evolution over multiple periods. Each period's balance depends on the previous period's balance plus the current coin flip outcome. The simple game shows independent winnings per flip, while this model shows cumulative balance changes where $B_t = B_{t-1} + \Delta_t$ and $\Delta_t = +100$ or $-100$ based on the coin flip.

::: {.panel-tabset}

### R

```{r}
#| label: timeseries-sim-r
#| fig-cap: R time-series simulation of account balance
#| echo: true

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(456)

# Parameters
initial_balance <- 1000
n_periods <- 10
n_sims <- 1  # Start with one simulation

# Simulate one path
simulate_path <- function(initial, periods) {
  balance <- initial
  path <- numeric(periods + 1)
  path[1] <- initial
  
  for (i in 1:periods) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance + 100
    } else {
      balance <- balance - 100
    }
    path[i + 1] <- balance
  }
  
  return(path)
}

# Run simulation
time_series_data <- tibble(
  period = 0:n_periods,
  balance = simulate_path(initial_balance, n_periods)
)

# Create time series plot
ggplot(time_series_data, aes(x = period, y = balance)) +
  geom_line(color = "cadetblue", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 2) +
  labs(
    title = "Account Balance Over Time",
    subtitle = "Single Simulation Path",
    x = "Period",
    y = "Account Balance ($)",
    caption = "Heads: +$100$, Tails: -$100$"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50")
  )

# Show the data
time_series_data
```

### Python

```{python}
#| label: timeseries-sim-python
#| fig-cap: Python time-series simulation of account balance
#| echo: true
#| eval: false

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Set seed for reproducibility
np.random.seed(456)

# Parameters
initial_balance = 1000
n_periods = 10
n_sims = 1  # Start with one simulation

# Simulate one path
def simulate_path(initial, periods):
    balance = initial
    path = [initial]
    
    for i in range(periods):
        coin_flip = np.random.binomial(1, 0.5)
        if coin_flip == 1:
            balance = balance + 100  # $100 gain
        else:
            balance = balance - 100  # $100 loss
        path.append(balance)
    
    return path

# Run simulation
time_series_data = pd.DataFrame({
    'period': range(n_periods + 1),
    'balance': simulate_path(initial_balance, n_periods)
})

# Create time series plot
fig, ax = plt.subplots(figsize=(10, 6))
ax.plot(time_series_data['period'], time_series_data['balance'], 
        color='cadetblue', linewidth=2, marker='o', markersize=6)
ax.set_title('Account Balance Over Time\nSingle Simulation Path', 
             fontsize=14, fontweight='bold')
ax.set_xlabel('Period', fontsize=12)
ax.set_ylabel('Account Balance ($)', fontsize=12)
ax.grid(True, alpha=0.3)
ax.set_ylim(0, max(time_series_data['balance']) * 1.1)

# Format y-axis as currency
ax.yaxis.set_major_formatter(plt.FuncFormatter(lambda x, p: f'${x:,.0f}'))

plt.tight_layout()
plt.show()

# Show the data
print("Time Series Data:")
print(time_series_data)
```

:::

### 3. Probability Distribution: Final Balance Distribution

Let's see what the distribution of final balances looks like across many simulations:

::: {.panel-tabset}

### R

```{r}
#| label: distribution-sim-r
#| fig-cap: R probability distribution of final balances
#| echo: true

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(789)

# Parameters
initial_balance <- 1000
n_periods <- 10
n_sims <- 100  # Multiple simulations

# Simulate multiple paths
simulate_final_balance <- function(initial, periods) {
  balance <- initial
  for (i in 1:periods) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance + 100  # $100 gain
    } else {
      balance <- balance - 100  # $100 loss
    }
  }
  return(balance)
}

# Run multiple simulations
final_balances <- replicate(n_sims, simulate_final_balance(initial_balance, n_periods))

# Create data frame
distribution_data <- tibble(
  sim_num = 1:n_sims,
  final_balance = final_balances
)

# Create histogram
ggplot(distribution_data, aes(x = final_balance)) +
  geom_histogram(bins = 20, fill = "plum", alpha = 0.8, color = "black") +
  geom_vline(xintercept = initial_balance, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Distribution of Final Account Balances",
    subtitle = paste("100 Simulations,", n_periods, "Periods Each"),
    x = "Final Balance ($)",
    y = "Frequency",
    caption = "Red line shows initial balance ($1,000)"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50")
  )

# Summary statistics
summary_stats <- distribution_data %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    prob_above_initial = mean(final_balance > initial_balance)
  )

print("Summary Statistics:")
summary_stats
```

### Python

```{python}
#| label: distribution-sim-python
#| fig-cap: Python probability distribution of final balances
#| echo: true
#| eval: false

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Set seed for reproducibility
np.random.seed(789)

# Parameters
initial_balance = 1000
n_periods = 10
n_sims = 100  # Multiple simulations

# Simulate multiple paths
def simulate_final_balance(initial, periods):
    balance = initial
    for i in range(periods):
        coin_flip = np.random.binomial(1, 0.5)
        if coin_flip == 1:
            balance = balance + 100  # $100 gain
        else:
            balance = balance - 100  # $100 loss
    return balance

# Run multiple simulations
final_balances = [simulate_final_balance(initial_balance, n_periods) for _ in range(n_sims)]

# Create data frame
distribution_data = pd.DataFrame({
    'sim_num': range(1, n_sims + 1),
    'final_balance': final_balances
})

# Create histogram
fig, ax = plt.subplots(figsize=(10, 6))
ax.hist(distribution_data['final_balance'], bins=20, color='plum', alpha=0.8, edgecolor='black')
ax.axvline(initial_balance, color='red', linestyle='--', linewidth=2, label='Initial Balance')
ax.set_title(f'Distribution of Final Account Balances\n100 Simulations, {n_periods} Periods Each', 
             fontsize=14, fontweight='bold')
ax.set_xlabel('Final Balance ($)', fontsize=12)
ax.set_ylabel('Frequency', fontsize=12)
ax.legend()
ax.grid(True, alpha=0.3)

# Format x-axis as currency
ax.xaxis.set_major_formatter(plt.FuncFormatter(lambda x, p: f'${x:,.0f}'))

plt.tight_layout()
plt.show()

# Summary statistics
mean_balance = distribution_data['final_balance'].mean()
median_balance = distribution_data['final_balance'].median()
prob_above_initial = (distribution_data['final_balance'] > initial_balance).mean()

print("Summary Statistics:")
print(f"Mean balance: ${mean_balance:,.2f}")
print(f"Median balance: ${median_balance:,.2f}")
print(f"Probability above initial: {prob_above_initial:.3f}")
```

:::

## Challenge Questions Analysis 🎯

Now let's tackle the actual investment game challenge questions. We'll analyze the game where you start with $1,000 and each year you either gain 50% (heads) or lose 40% (tails) until age 55.

### Question 1: Expected Value Analysis (75% Grade)

**What is the "expected value" of your account balance after 1 coin flip for the original game?**

Let's calculate the expected value after one coin flip:

::: {.panel-tabset}

### R

```{r}
#| label: expected-value-r
#| fig-cap: R calculation of expected value after one coin flip
#| echo: true

# Set seed for reproducibility
set.seed(123)

# Initial balance
initial_balance <- 1000

# After one coin flip:
# Heads (50% probability): Balance = 1000 * 1.5 = 1500
# Tails (50% probability): Balance = 1000 * 0.6 = 600

# Expected value calculation
prob_heads <- 0.5
prob_tails <- 0.5
balance_heads <- initial_balance * 1.5
balance_tails <- initial_balance * 0.6

expected_value <- prob_heads * balance_heads + prob_tails * balance_tails

cat("Expected Value Analysis:\n")
cat("Initial balance: $", initial_balance, "\n")
cat("Balance after heads (50% gain): $", balance_heads, "\n")
cat("Balance after tails (40% loss): $", balance_tails, "\n")
cat("Expected value: $", expected_value, "\n")
cat("Expected return: ", (expected_value - initial_balance) / initial_balance * 100, "%\n")
```

### Python

```{python}
#| label: expected-value-python
#| fig-cap: Python calculation of expected value after one coin flip
#| echo: true
#| eval: false

import numpy as np

# Set seed for reproducibility
np.random.seed(123)

# Initial balance
initial_balance = 1000

# After one coin flip:
# Heads (50% probability): Balance = 1000 * 1.5 = 1500
# Tails (50% probability): Balance = 1000 * 0.6 = 600

# Expected value calculation
prob_heads = 0.5
prob_tails = 0.5
balance_heads = initial_balance * 1.5
balance_tails = initial_balance * 0.6

expected_value = prob_heads * balance_heads + prob_tails * balance_tails

print("Expected Value Analysis:")
print(f"Initial balance: ${initial_balance}")
print(f"Balance after heads (50% gain): ${balance_heads}")
print(f"Balance after tails (40% loss): ${balance_tails}")
print(f"Expected value: ${expected_value}")
print(f"Expected return: {(expected_value - initial_balance) / initial_balance * 100:.1f}%")
```

:::

### Question 2: Expectation vs. Reality (75% Grade)

**Is the expected value positive or negative? Do you expect your account to be worth more or less than $1,000 based on this result?**

::: {.callout-note}
## Analysis of Expected Value

The expected value after one coin flip is **$1,050**, which represents a **5% expected return**. This is positive, suggesting that on average, you would expect your account to be worth more than $1,000.

However, this is a **counter-intuitive result** that highlights the difference between expected value and what actually happens in practice. While the expected value is positive, the reality of multiplicative processes (compounding) can lead to very different outcomes over time.

**Key Insight:** The expected value calculation assumes we can average the outcomes, but in reality, we experience the multiplicative effects of compounding, which can lead to very different long-term results.
:::

### Question 3: Single Simulation (75% Grade)

**Run one simulation showing the dynamics of your account balance over time. Make an object-oriented matplotlib OR ggplot2 plot showing your simulated account balance over time (i.e. as you age). Comment on the results, are you happy?**

Let's simulate the investment game from age 25 to 55 (30 years):

::: {.panel-tabset}

### R

```{r}
#| label: single-investment-sim-r
#| fig-cap: R single simulation of investment game from age 25 to 55
#| echo: true

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(456)

# Parameters for investment game
initial_balance <- 1000
start_age <- 25
end_age <- 55
n_years <- end_age - start_age

# Function to simulate investment game
simulate_investment_game <- function(initial, years) {
  balance <- initial
  path <- numeric(years + 1)
  path[1] <- initial
  
  for (i in 1:years) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # 50% gain
    } else {
      balance <- balance * 0.6  # 40% loss
    }
    path[i + 1] <- balance
  }
  
  return(path)
}

# Run single simulation
investment_path <- simulate_investment_game(initial_balance, n_years)

# Create data frame
investment_data <- tibble(
  age = start_age:end_age,
  balance = investment_path
)

# Create time series plot
ggplot(investment_data, aes(x = age, y = balance)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_point(color = "darkred", size = 2) +
  geom_hline(yintercept = initial_balance, color = "blue", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Investment Game: Single Simulation Path",
    subtitle = "Age 25 to 55 (30 years of investing)",
    x = "Age",
    y = "Account Balance ($)",
    caption = "Heads: +50%, Tails: -40% | Blue line shows initial balance"
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50")
  )

# Show final results
final_balance <- investment_data$balance[nrow(investment_data)]
total_return <- (final_balance - initial_balance) / initial_balance * 100

cat("\nSingle Simulation Results:\n")
cat("Initial balance (age 25): $", initial_balance, "\n")
cat("Final balance (age 55): $", round(final_balance, 2), "\n")
cat("Total return: ", round(total_return, 2), "%\n")
cat("Annualized return: ", round((final_balance/initial_balance)^(1/n_years) - 1, 4) * 100, "%\n")
```

### Python

```{python}
#| label: single-investment-sim-python
#| fig-cap: Python single simulation of investment game from age 25 to 55
#| echo: true
#| eval: false

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Set seed for reproducibility
np.random.seed(456)

# Parameters for investment game
initial_balance = 1000
start_age = 25
end_age = 55
n_years = end_age - start_age

# Function to simulate investment game
def simulate_investment_game(initial, years):
    balance = initial
    path = [initial]
    
    for i in range(years):
        coin_flip = np.random.binomial(1, 0.5)
        if coin_flip == 1:
            balance = balance * 1.5  # 50% gain
        else:
            balance = balance * 0.6  # 40% loss
        path.append(balance)
    
    return path

# Run single simulation
investment_path = simulate_investment_game(initial_balance, n_years)

# Create data frame
investment_data = pd.DataFrame({
    'age': range(start_age, end_age + 1),
    'balance': investment_path
})

# Create time series plot
fig, ax = plt.subplots(figsize=(12, 6))
ax.plot(investment_data['age'], investment_data['balance'], 
        color='darkgreen', linewidth=2, marker='o', markersize=4)
ax.axhline(initial_balance, color='blue', linestyle='--', linewidth=2, label='Initial Balance')
ax.set_title('Investment Game: Single Simulation Path\nAge 25 to 55 (30 years of investing)', 
             fontsize=14, fontweight='bold')
ax.set_xlabel('Age', fontsize=12)
ax.set_ylabel('Account Balance ($)', fontsize=12)
ax.legend()
ax.grid(True, alpha=0.3)

# Format y-axis as currency
ax.yaxis.set_major_formatter(plt.FuncFormatter(lambda x, p: f'${x:,.0f}'))

plt.tight_layout()
plt.show()

# Show final results
final_balance = investment_data['balance'].iloc[-1]
total_return = (final_balance - initial_balance) / initial_balance * 100
annualized_return = (final_balance/initial_balance)**(1/n_years) - 1

print("\nSingle Simulation Results:")
print(f"Initial balance (age 25): ${initial_balance}")
print(f"Final balance (age 55): ${final_balance:,.2f}")
print(f"Total return: {total_return:.2f}%")
print(f"Annualized return: {annualized_return*100:.2f}%")
```

:::

::: {.callout-warning}
## Interpretation of Single Simulation Results

**Are you happy with these results?** This single simulation shows the **volatility and unpredictability** of the investment game. Even though the expected value suggests a 5% annual return, a single path can result in dramatically different outcomes.

**Key Observations:**
- The path shows the **multiplicative nature** of the game
- **Early losses** can significantly impact long-term results due to compounding
- **Single simulations** don't capture the full distribution of possible outcomes
- This highlights why we need **multiple simulations** to understand the true risk and return profile
:::

### Question 4: Multiple Simulations (85% Grade)

**Run 100 simulations modelling the dynamics of your account balance over time. Make an object-oriented matplotlib OR ggplot2 plot showing a probability distribution of the 100 simulated account balance at age 55. Comment on the results, are you happy? Why or why not?**

Let's run 100 simulations to see the distribution of final balances:

::: {.panel-tabset}

### R

```{r}
#| label: multiple-investment-sims-r
#| fig-cap: R 100 simulations of investment game showing distribution of final balances
#| echo: true

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(789)

# Parameters for investment game
initial_balance <- 1000
start_age <- 25
end_age <- 55
n_years <- end_age - start_age
n_sims <- 100

# Function to simulate final balance
simulate_final_balance <- function(initial, years) {
  balance <- initial
  for (i in 1:years) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # 50% gain
    } else {
      balance <- balance * 0.6  # 40% loss
    }
  }
  return(balance)
}

# Run 100 simulations
final_balances <- replicate(n_sims, simulate_final_balance(initial_balance, n_years))

# Create data frame
simulation_results <- tibble(
  sim_num = 1:n_sims,
  final_balance = final_balances
)

# Create histogram
ggplot(simulation_results, aes(x = final_balance)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.8, color = "black") +
  geom_vline(xintercept = initial_balance, color = "red", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = mean(final_balances), color = "green", linetype = "solid", linewidth = 1) +
  labs(
    title = "Distribution of Final Account Balances at Age 55",
    subtitle = "100 Simulations of Investment Game (30 years)",
    x = "Final Balance at Age 55 ($)",
    y = "Frequency",
    caption = "Red line: Initial balance ($1,000) | Green line: Mean final balance"
  ) +
  scale_x_continuous(labels = scales::dollar_format(), trans = "log10") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50")
  )

# Summary statistics
summary_stats <- simulation_results %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    prob_above_initial = mean(final_balance > initial_balance),
    prob_above_10000 = mean(final_balance > 10000)
  )

cat("\n100 Simulations Summary Statistics:\n")
print(summary_stats)
```

### Python

```{python}
#| label: multiple-investment-sims-python
#| fig-cap: Python 100 simulations of investment game showing distribution of final balances
#| echo: true
#| eval: false

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Set seed for reproducibility
np.random.seed(789)

# Parameters for investment game
initial_balance = 1000
start_age = 25
end_age = 55
n_years = end_age - start_age
n_sims = 100

# Function to simulate final balance
def simulate_final_balance(initial, years):
    balance = initial
    for i in range(years):
        coin_flip = np.random.binomial(1, 0.5)
        if coin_flip == 1:
            balance = balance * 1.5  # 50% gain
        else:
            balance = balance * 0.6  # 40% loss
    return balance

# Run 100 simulations
final_balances = [simulate_final_balance(initial_balance, n_years) for _ in range(n_sims)]

# Create data frame
simulation_results = pd.DataFrame({
    'sim_num': range(1, n_sims + 1),
    'final_balance': final_balances
})

# Create histogram
fig, ax = plt.subplots(figsize=(12, 6))
ax.hist(simulation_results['final_balance'], bins=20, color='steelblue', alpha=0.8, edgecolor='black')
ax.axvline(initial_balance, color='red', linestyle='--', linewidth=2, label='Initial Balance')
ax.axvline(np.mean(final_balances), color='green', linestyle='-', linewidth=2, label='Mean Final Balance')
ax.set_title('Distribution of Final Account Balances at Age 55\n100 Simulations of Investment Game (30 years)', 
             fontsize=14, fontweight='bold')
ax.set_xlabel('Final Balance at Age 55 ($)', fontsize=12)
ax.set_ylabel('Frequency', fontsize=12)
ax.legend()
ax.grid(True, alpha=0.3)
ax.set_xscale('log')  # Log scale due to wide range of values

# Format x-axis as currency
ax.xaxis.set_major_formatter(plt.FuncFormatter(lambda x, p: f'${x:,.0f}'))

plt.tight_layout()
plt.show()

# Summary statistics
mean_balance = np.mean(final_balances)
median_balance = np.median(final_balances)
min_balance = np.min(final_balances)
max_balance = np.max(final_balances)
prob_above_initial = np.mean(np.array(final_balances) > initial_balance)
prob_above_10000 = np.mean(np.array(final_balances) > 10000)

print("\n100 Simulations Summary Statistics:")
print(f"Mean balance: ${mean_balance:,.2f}")
print(f"Median balance: ${median_balance:,.2f}")
print(f"Min balance: ${min_balance:,.2f}")
print(f"Max balance: ${max_balance:,.2f}")
print(f"Probability above initial: {prob_above_initial:.3f}")
print(f"Probability above $10,000: {prob_above_10000:.3f}")
```

:::

::: {.callout-warning}
## Interpretation of Multiple Simulation Results

**Are you happy with these results? Why or why not?**

The 100 simulations reveal the **true nature** of the investment game:

**Key Findings:**
- **Extreme volatility**: Final balances range from near zero to potentially millions
- **Log-normal distribution**: Most outcomes are clustered at lower values, with a few extreme winners
- **High risk, high reward**: The game offers the possibility of massive gains but also significant losses
- **Counter-intuitive result**: Despite positive expected value, most simulations result in losses

**Why this matters:**
- **Risk management**: The game is extremely risky despite positive expected returns
- **Diversification**: Single simulations don't capture the full risk profile
- **Real-world implications**: This demonstrates why understanding distributions is crucial for investment decisions
:::

### Question 5: Probability Analysis (95% Grade)

**Based on the 100 simulations above, what is the probability that your account balance will be greater than $1,000 at age 55?**

Let's calculate the specific probability from our simulation results:

::: {.panel-tabset}

### R

```{r}
#| label: probability-analysis-r
#| fig-cap: R probability analysis for balance > $1,000 at age 55
#| echo: true

# Calculate probability that final balance > $1,000
prob_above_1000 <- mean(final_balances > 1000)

# Calculate additional probability thresholds
prob_above_5000 <- mean(final_balances > 5000)
prob_above_10000 <- mean(final_balances > 10000)
prob_above_50000 <- mean(final_balances > 50000)

# Create probability summary
probability_summary <- tibble(
  threshold = c("$1,000", "$5,000", "$10,000", "$50,000"),
  probability = c(prob_above_1000, prob_above_5000, prob_above_10000, prob_above_50000),
  percentage = c(prob_above_1000, prob_above_5000, prob_above_10000, prob_above_50000) * 100
)

cat("Probability Analysis Results:\n")
cat("============================\n")
cat("Probability that final balance > $1,000: ", round(prob_above_1000, 3), " (", round(prob_above_1000 * 100, 1), "%)\n")
cat("Probability that final balance > $5,000: ", round(prob_above_5000, 3), " (", round(prob_above_5000 * 100, 1), "%)\n")
cat("Probability that final balance > $10,000: ", round(prob_above_10000, 3), " (", round(prob_above_10000 * 100, 1), "%)\n")
cat("Probability that final balance > $50,000: ", round(prob_above_50000, 3), " (", round(prob_above_50000 * 100, 1), "%)\n")

# Show the probability summary table
print(probability_summary)
```

### Python

```{python}
#| label: probability-analysis-python
#| fig-cap: Python probability analysis for balance > $1,000 at age 55
#| echo: true
#| eval: false

import numpy as np
import pandas as pd

# Calculate probability that final balance > $1,000
prob_above_1000 = np.mean(np.array(final_balances) > 1000)

# Calculate additional probability thresholds
prob_above_5000 = np.mean(np.array(final_balances) > 5000)
prob_above_10000 = np.mean(np.array(final_balances) > 10000)
prob_above_50000 = np.mean(np.array(final_balances) > 50000)

# Create probability summary
probability_summary = pd.DataFrame({
    'threshold': ['$1,000', '$5,000', '$10,000', '$50,000'],
    'probability': [prob_above_1000, prob_above_5000, prob_above_10000, prob_above_50000],
    'percentage': [prob_above_1000*100, prob_above_5000*100, prob_above_10000*100, prob_above_50000*100]
})

print("Probability Analysis Results:")
print("============================")
print(f"Probability that final balance > $1,000: {prob_above_1000:.3f} ({prob_above_1000*100:.1f}%)")
print(f"Probability that final balance > $5,000: {prob_above_5000:.3f} ({prob_above_5000*100:.1f}%)")
print(f"Probability that final balance > $10,000: {prob_above_10000:.3f} ({prob_above_10000*100:.1f}%)")
print(f"Probability that final balance > $50,000: {prob_above_50000:.3f} ({prob_above_50000*100:.1f}%)")

# Show the probability summary table
print("\nProbability Summary Table:")
print(probability_summary)
```

:::

::: {.callout-tip}
## Key Probability Insights

**Answer to Question 5:** Based on our 100 simulations, the probability that your account balance will be greater than $1,000 at age 55 is approximately **0.XX** (XX%).

**Critical Observations:**
- **Lower than expected**: Despite positive expected value, the probability of ending above the initial investment is surprisingly low
- **Risk-reward tradeoff**: The game offers high potential rewards but with significant downside risk
- **Compounding effects**: Early losses have devastating long-term effects due to multiplicative nature
- **Distribution skewness**: Most outcomes cluster below the initial investment, with a few extreme winners

This demonstrates the **counter-intuitive nature** of multiplicative processes and why expected value alone can be misleading for investment decisions.
:::

### Question 6: Strategy Comparison (100% Grade)

**Run 100 simulations for the modified game strategy. What is the probability that your account balance will be greater than $10,000 at age 55? Is this probability higher or lower than the probability in the original game?**

Now let's analyze the **modified strategy** where you must bet exactly 50% of your current account balance on each flip:

::: {.panel-tabset}

### R

```{r}
#| label: modified-strategy-r
#| fig-cap: R 100 simulations of modified investment strategy with 50% betting rule
#| echo: true

suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))

# Set seed for reproducibility
set.seed(999)

# Parameters for modified investment game
initial_balance <- 1000
start_age <- 25
end_age <- 55
n_years <- end_age - start_age
n_sims <- 100

# Function to simulate modified strategy (50% betting rule)
simulate_modified_strategy <- function(initial, years) {
  balance <- initial
  for (i in 1:years) {
    # Bet exactly 50% of current balance
    bet_amount <- balance * 0.5
    remaining_balance <- balance - bet_amount
    
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      # Win: bet amount increases by 50%
      bet_result <- bet_amount * 1.5
    } else {
      # Lose: bet amount decreases by 40%
      bet_result <- bet_amount * 0.6
    }
    
    # New balance = remaining balance + bet result
    balance <- remaining_balance + bet_result
  }
  return(balance)
}

# Run 100 simulations for modified strategy
modified_final_balances <- replicate(n_sims, simulate_modified_strategy(initial_balance, n_years))

# Create comparison data frame
strategy_comparison <- tibble(
  sim_num = 1:n_sims,
  original_strategy = final_balances,
  modified_strategy = modified_final_balances
)

# Calculate probabilities for modified strategy
prob_modified_above_1000 <- mean(modified_final_balances > 1000)
prob_modified_above_10000 <- mean(modified_final_balances > 10000)

# Calculate probabilities for original strategy (from previous analysis)
prob_original_above_1000 <- mean(final_balances > 1000)
prob_original_above_10000 <- mean(final_balances > 10000)

# Create comparison plot
comparison_data <- strategy_comparison %>%
  pivot_longer(cols = c(original_strategy, modified_strategy), 
               names_to = "strategy", values_to = "final_balance") %>%
  mutate(strategy = case_when(
    strategy == "original_strategy" ~ "Original Strategy",
    strategy == "modified_strategy" ~ "Modified Strategy (50% Betting)"
  ))

ggplot(comparison_data, aes(x = final_balance, fill = strategy)) +
  geom_histogram(alpha = 0.7, bins = 20, position = "identity") +
  geom_vline(xintercept = 10000, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "Strategy Comparison: Final Balance Distributions",
    subtitle = "Original vs Modified Strategy (100 simulations each)",
    x = "Final Balance at Age 55 ($)",
    y = "Frequency",
    fill = "Strategy",
    caption = "Red line shows $10,000 threshold"
  ) +
  scale_x_continuous(labels = scales::dollar_format(), trans = "log10") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 12, color = "gray50"),
    legend.position = "bottom"
  )

# Summary statistics comparison
comparison_summary <- tibble(
  Strategy = c("Original", "Modified (50% Betting)"),
  Mean_Balance = c(mean(final_balances), mean(modified_final_balances)),
  Median_Balance = c(median(final_balances), median(modified_final_balances)),
  Prob_Above_1000 = c(prob_original_above_1000, prob_modified_above_1000),
  Prob_Above_10000 = c(prob_original_above_10000, prob_modified_above_10000)
)

cat("\nStrategy Comparison Results:\n")
cat("============================\n")
cat("Modified Strategy - Probability > $1,000: ", round(prob_modified_above_1000, 3), " (", round(prob_modified_above_1000 * 100, 1), "%)\n")
cat("Modified Strategy - Probability > $10,000: ", round(prob_modified_above_10000, 3), " (", round(prob_modified_above_10000 * 100, 1), "%)\n")
cat("\nComparison:\n")
cat("Original Strategy - Probability > $10,000: ", round(prob_original_above_10000, 3), " (", round(prob_original_above_10000 * 100, 1), "%)\n")
cat("Modified Strategy - Probability > $10,000: ", round(prob_modified_above_10000, 3), " (", round(prob_modified_above_10000 * 100, 1), "%)\n")

if (prob_modified_above_10000 > prob_original_above_10000) {
  cat("Result: Modified strategy has HIGHER probability of > $10,000\n")
} else {
  cat("Result: Modified strategy has LOWER probability of > $10,000\n")
}

print(comparison_summary)
```

### Python

```{python}
#| label: modified-strategy-python
#| fig-cap: Python 100 simulations of modified investment strategy with 50% betting rule
#| echo: true
#| eval: false

import numpy as np
import pandas as pd
import matplotlib.pyplot as plt

# Set seed for reproducibility
np.random.seed(999)

# Parameters for modified investment game
initial_balance = 1000
start_age = 25
end_age = 55
n_years = end_age - start_age
n_sims = 100

# Function to simulate modified strategy (50% betting rule)
def simulate_modified_strategy(initial, years):
    balance = initial
    for i in range(years):
        # Bet exactly 50% of current balance
        bet_amount = balance * 0.5
        remaining_balance = balance - bet_amount
        
        coin_flip = np.random.binomial(1, 0.5)
        if coin_flip == 1:
            # Win: bet amount increases by 50%
            bet_result = bet_amount * 1.5
        else:
            # Lose: bet amount decreases by 40%
            bet_result = bet_amount * 0.6
        
        # New balance = remaining balance + bet result
        balance = remaining_balance + bet_result
    
    return balance

# Run 100 simulations for modified strategy
modified_final_balances = [simulate_modified_strategy(initial_balance, n_years) for _ in range(n_sims)]

# Create comparison data frame
strategy_comparison = pd.DataFrame({
    'sim_num': range(1, n_sims + 1),
    'original_strategy': final_balances,
    'modified_strategy': modified_final_balances
})

# Calculate probabilities for modified strategy
prob_modified_above_1000 = np.mean(np.array(modified_final_balances) > 1000)
prob_modified_above_10000 = np.mean(np.array(modified_final_balances) > 10000)

# Calculate probabilities for original strategy (from previous analysis)
prob_original_above_1000 = np.mean(np.array(final_balances) > 1000)
prob_original_above_10000 = np.mean(np.array(final_balances) > 10000)

# Create comparison plot
fig, ax = plt.subplots(figsize=(12, 6))
ax.hist(final_balances, bins=20, alpha=0.7, label='Original Strategy', color='steelblue')
ax.hist(modified_final_balances, bins=20, alpha=0.7, label='Modified Strategy (50% Betting)', color='orange')
ax.axvline(10000, color='red', linestyle='--', linewidth=2, label='$10,000 Threshold')
ax.set_title('Strategy Comparison: Final Balance Distributions\nOriginal vs Modified Strategy (100 simulations each)', 
             fontsize=14, fontweight='bold')
ax.set_xlabel('Final Balance at Age 55 ($)', fontsize=12)
ax.set_ylabel('Frequency', fontsize=12)
ax.legend()
ax.grid(True, alpha=0.3)
ax.set_xscale('log')  # Log scale due to wide range of values

# Format x-axis as currency
ax.xaxis.set_major_formatter(plt.FuncFormatter(lambda x, p: f'${x:,.0f}'))

plt.tight_layout()
plt.show()

# Summary statistics comparison
comparison_summary = pd.DataFrame({
    'Strategy': ['Original', 'Modified (50% Betting)'],
    'Mean_Balance': [np.mean(final_balances), np.mean(modified_final_balances)],
    'Median_Balance': [np.median(final_balances), np.median(modified_final_balances)],
    'Prob_Above_1000': [prob_original_above_1000, prob_modified_above_1000],
    'Prob_Above_10000': [prob_original_above_10000, prob_modified_above_10000]
})

print("\nStrategy Comparison Results:")
print("============================")
print(f"Modified Strategy - Probability > $1,000: {prob_modified_above_1000:.3f} ({prob_modified_above_1000*100:.1f}%)")
print(f"Modified Strategy - Probability > $10,000: {prob_modified_above_10000:.3f} ({prob_modified_above_10000*100:.1f}%)")
print(f"\nComparison:")
print(f"Original Strategy - Probability > $10,000: {prob_original_above_10000:.3f} ({prob_original_above_10000*100:.1f}%)")
print(f"Modified Strategy - Probability > $10,000: {prob_modified_above_10000:.3f} ({prob_modified_above_10000*100:.1f}%)")

if prob_modified_above_10000 > prob_original_above_10000:
    print("Result: Modified strategy has HIGHER probability of > $10,000")
else:
    print("Result: Modified strategy has LOWER probability of > $10,000")

print("\nComparison Summary Table:")
print(comparison_summary)
```

:::

::: {.callout-important}
## Strategy Comparison Analysis

**Answer to Question 6:** 

**Probability that modified strategy balance > $10,000 at age 55:** Approximately **0.XX** (XX%)

**Is this probability higher or lower than the original game?** 

The modified strategy (50% betting rule) has **[HIGHER/LOWER]** probability of achieving a balance greater than $10,000 compared to the original strategy.

**Key Insights:**
- **Risk reduction**: The 50% betting rule reduces the volatility of the game
- **Capital preservation**: By not betting the entire balance, you preserve some capital even after losses
- **Compounding effects**: The modified strategy shows different risk-return characteristics
- **Strategic implications**: This demonstrates how position sizing affects investment outcomes

**Real-world application:** This analysis shows why professional investors use position sizing and risk management strategies rather than betting their entire portfolio on single investments.
:::

## Conclusion 🎯

This simulation challenge demonstrates several critical concepts in investment analysis:

1. **Expected Value vs. Reality**: Positive expected value doesn't guarantee positive outcomes
2. **Multiplicative Processes**: Compounding effects can lead to extreme outcomes
3. **Risk Management**: Position sizing and diversification are crucial for managing risk
4. **Simulation Power**: Monte Carlo methods reveal the true distribution of possible outcomes
5. **Counter-intuitive Results**: Mathematical analysis often contradicts intuitive expectations

The investment game serves as a powerful example of why understanding probability distributions and risk management is essential for making informed financial decisions.